"Advanced" Data Manipulation
========================================================

In progress.

TODO:
* dplyr for data manipulation

Additional functions to cover:

* metrumrg package:

  * check.subjects
  * locf
  * metaMerge
  * metaSub
  * resample.data.frame
  

  

Example:

split theoph by id, get max conc value and bind it to new column, do.call and rbind to return to same structure

```{r }
Theoph_ids <- split(Theoph, with(Theoph, Subject))
cmax <- lapply(Theoph_ids, function(x) within(x, cmax <- max(conc)) )
head(cmax)
cmax_df <- do.call(rbind, cmax)
head(cmax_df)
rownames(cmax_df) <- NULL
head(cmax_df)
```

##Passing other arguments into apply-style calls

`cmax <- lapply(Theoph_ids, function(x, y) within(x, cmax <- max(y)), "conc" )`

TODO: add str-split and some more data transformation stuff

TODO: clean up data transformation stuff
subjid <- "site001-AB-01"

substr(subjid, 12, 13)
?strsplit
site <- strsplit(subjid, "-")

subTheoph <- Theoph[Theoph$Subj ==1,]
subTheoph$subjid <- "site001-AB-01"
within(subTheoph, site <- unlist(strsplit(subjid, "-"))[1])

TODO: add dplyr section

#### moved from intro data manipulation

### Additional manipulations 

### Ordering and expand.grid()
The concept of keys can help us think about the structure of the dataset when we deal with ordering.

The `order` function in R allows us to easily sort data in ascending or descending order. 

If we wanted to sort a dataset by ID and Time we could do so via:

`df <- df[order(df$ID, df$Time),]`

It is a good habit to sort raw data (especially if its eventually going into phoenix or nonmem)

The way `order` works with multiple arguments is it sorts starting with the first argument, each time it runs into a 'tie' it looks for subsequent arguments for how to break the 'tie'

For example, if you have 5 time observations associated with an ID, if you do:
`order(df$ID, df$TIME)` it will start sorting by `ID`, notice it has multiple results for the same ID value, it will then use `TIME` to continue the sort.

`order` is quite powerful with the `expand.grid` function. 



To quickly note - columns and rows can be removed simply by calling their index and assigning it to NULL

`Theoph$Subject <- NULL` or `Theoph[Theoph$Subject ==1,] <- NULL`

`with` - 'opens up' the input within the function to allow us access columns similar to `attach()` but only in the function's temporary environment. Can be very useful for with modeling.

```{r }
with(Theoph, summary(conc))
# same as
summary(Theoph$conc)
```

`within` - allows us access inside a data structure to do things *within* it

``` {r }
id_time <- data.frame(expand.grid(ID = 1:3, TIME = 0:10))
head(id_time)
id_time <- id_time[order(id_time$ID, id_time$TIME),]
head(id_time)
df <- within(id_time, {
  DV <- 0
  AMT <- 0
  DOSE <- 150
  MDV <- 0
})
head(df)
```

ie will have column order MDV, DOSE, AMT, DV add by column from bottom up going outward

**notice the structure of a `within` call, there are no `,`s inside the brackets when creating multiple columns

Let's try to create a simulation data frame

```{r }
df <- data.frame(ID = 1:5)
 df <- within(df, {
   UCTX <- 0
   # initialize amount to NA so will be read in as blank to phoenix
   AMT <- NA
   # initialize doses
   DOSE <- numeric()
   ADDL <- NA
   II <- NA
 })
str(df)
```

Check out the differences between AMT/ADDL/II columns and DOSE

`within` can also be used for logical subsetting in a concisely written format:

```{r }
Theoph$MDV <- 1
Theoph <- within(Theoph, MDV[Subject == 2 & conc == 0] <- 0)
```

## Assignments
- using `expand.grid` and `order` create an `id_time` dataset (dataframe)
- within the dataset add another column `Measurement` that is a set of noise generated with `rnorm` or `runif`
- use `with` to subset Theoph for ID's 1 and 2 only
- generate a new conc column for Theoph

