Apply Functions
========================================================

TODO: flesh out and add more concrete PMTRX examples

There are a number of `apply` commands in the R family. Each has different way of handling data

Additional information can be found in threads on [stackoverflow](http://stackoverflow.com/questions/3505701/r-grouping-functions-sapply-vs-lapply-vs-apply-vs-tapply-vs-by-vs-aggrega)

Note - the apply-style commands are *not* particularly faster than a *well-written* loop. The benefit is their one-line nature. [Read some more here](http://stackoverflow.com/questions/2275896/is-rs-apply-family-more-than-syntactic-sugar?rq=1)

Here is a brief overview:

TODO: test tag here

## apply

Used to evaluate a function over the margins of a matrix/array. It can apply a function by dimension (row/col).  

`apply` takes the following arguments: 
* `X` - input
* `MARGIN` - to distinguish by row or column. `1` - row, `2` - column
* `FUN` - function to be applied
* `...` is for other arguments to be passed into fun

For example to apply the function `range` across all columns in `Theoph`:

```{r }
apply(Theoph, 2, range)
str(apply(Theoph, 2, range))
```

Notice:
- the 'range' function does not have '()' around it.
- The results are all characters - since `apply` goes over by dimension and needs a matrix/array, the whole data structure is coerced to the highest level. Since Subject is an ordered factor for `Theoph` the whole data-frame is coerced to a matrix of type `char` before the function is applied!

*tidbit* - apply is a good opportunity to use colMeans/colSums and rowMeans/rowSums - they are significantly faster, especially for large data structures.


## lapply

Works to apply a function to each element in the list and *returns a list back*

`lapply` takes the following arguments:
* `X` - input
* `FUN` - function to be applied
* `...` is for other arguments to be passed into fun

```{r }
lapply(Theoph, range)
```

Given that a dataframe is a list with dimensions - lapply can easily step throuh the columns and apply your function and returns a list.

## sapply

sapply is similar to lapply, the biggest difference is it attempts to simplify the result.

```{r }
lapply(Theoph, range)
sapply(Theoph, range)
str(lapply(Theoph, range))
str(sapply(Theoph, range))
class(lapply(Theoph, range))
class(sapply(Theoph, range))
typeof(lapply(Theoph, range))
typeof(sapply(Theoph, range))
```

Generally, the "rules" for `sapply` can be thought-of as:
* Result is a list where every element is length 1 - returns a vector
* Result list where every elemtn is same length - returns a matrix
* Neither of above - returns list

## vapply

 preferred over sapply due to minor speed improvement and better consistency with return types. Can read more about it on [stack overflow here](http://stackoverflow.com/questions/12339650/why-is-vapply-safer-than-sapply)

## tapply

Used to apply functions to subsets of a vector. I will briefly give some detail, though my person preference is to use `plyr` commands or `split` when I'm dealing with subsets.

`tapply` takes the following arguments: 
* `X` - input
* `INDEX` is a factor or list of factors
* `FUN` - function to be applied
* `...` is for other arguments to be passed into fun
* `simplify` - logical for whether to simplify result or not

**note**: INDEX must be a factor or list of factors - if not any value passed in will be coerced to factor.

```{r }
with(Theoph, tapply(conc, Subject, mean))
```

Without using with:

```{r }
tapply(Theoph$conc, Theoph$Subject, mean)
```

If speed is of the essence - `tapply` does perform well - for some minor benchmarks it proved to be ~ 7x faster than plyr - though take that with a grain of salt as it was the difference between 0.19 and 1.3 seconds. Hadley Wickham is also coming out with his new `dplyr` package that should increase the speed of `plyr`-style commands substantially.

## reapply

 metrumrg function - similar to `tapply` but adds some additional functionality, including a `where` option to restrict calculations in domain of `x`.

```{r }
library(metrumrg)
with(Theoph,reapply(conc,Subject,mean,where=Time!=0,na.rm=TRUE))
```

