Control Structors 
========================================================

Testing Conditions

```
if
if, else if 
if, else if, ..., else
```

Have the structure

```
if(condition) {
    ## do stuff
} else if (conditition2) {
    ## do other stuff
    } else {
    ## do something in for other condition
}
```

You don't have to use `else` statments

```
if (condition1)
if (condition2)
if (condition3)
...
```

is perfectly valid.

* `ifelse` - specialized test with the format `ifelse(Test, Yes, no)`

* `for` looping structure

Loops can be constructed based on a specified vector length or by specific indices

```
for (i in 1:5) {
    store_results[i] <- do_something()
}
```

While you may be more familiar with the construct:

```
for(i in 1:length(x)) {
     results[i] <- do_something(x[i])
}
```

This is actually a "bad" habit that can run you into trouble with objects of length(0)

`seq_along` is a 'safer' option that has the exact same effect if you're starting from the first indice, with the added benefit of failing more gracefully.

* `while` - execute loop *while* tested condition is true. Often used if you need to evaluate until a certain condition is met. **danger**: can potentially result in infinite loops if not written properly or if the tested condition is never met.

```{r }
count <- 0
while(count < 10) {
    print(count)
    count <- count + 1
}
```

* `next` - skip iteration of loop

`next` can be used to skip iterations in a loop 

```{r }
for(i in 1:10) {
    if (i < 3) next
    print(i)
}
```

`break` break the execution of a loop

`return` - exit function


## Assignments
* within theoph add a new column called `ISEVEN` - for all subjucts with even ID numbers assign a value 1, for all odds assign value 0
* write a creatinine clearance calculator function - use it to calculate the CRCL for some simulated data in a vectorized manner
* 